\section{Matrices and linear systems} \label{sec:matrix}

\LAtt{3.2}

\LO{
\item Define and perform addition and multiplication operations on matrices,
\item Compute the determinant of a square matrix, and
\item Find eigenvalues and eigenvectors of a square matrix. 
}

This section is meant to summarize the parts of linear algebra that will be necessary in the process of developing and solving linear systems of differential equations. All of this information is covered in more detail in \Chapterref{linalg:appendix}, so you can find more information there. If you went through that chapter already, this section will serve as a review.

\subsection{Matrices and vectors}

Before we start talking about linear systems of ODEs, we need to
talk about matrices, so let us review these briefly.  A \emph{\myindex{matrix}}
is an $m
\times n$ array of numbers ($m$ rows and $n$ columns).  For example, we denote
a $3 \times 5$ matrix as follows
\begin{equation*}
A = 
\begin{bmatrix}
a_{11} & a_{12} & a_{13} & a_{14} & a_{15} \\
a_{21} & a_{22} & a_{23} & a_{24} & a_{25} \\
a_{31} & a_{32} & a_{33} & a_{34} & a_{35}
\end{bmatrix} .
\end{equation*}
The numbers $a_{ij}$ are called \emph{elements}\index{element of a matrix}
or \emph{entries}\index{entry of a matrix}.

By a \emph{\myindex{vector}} we usually mean a
\emph{\myindex{column vector}}, that is an $m \times 1$ matrix.
If we mean a \emph{\myindex{row vector}},
we will explicitly say so (a row vector is a $1 \times n$ matrix).
We usually denote
matrices by upper case letters and vectors by lower case letters with an
arrow such as $\vec{x}$ or $\vec{b}$.  By $\vec{0}$ we mean the vector
of all zeros.

We define some operations on matrices.  We 
want $1 \times 1$ matrices to really act like numbers, so our operations
have to be compatible with this viewpoint.

First, we can multiply\index{scalar multiplication} a matrix by
a \emph{\myindex{scalar}} (a number).
We simply multiply each entry in the matrix by the scalar.  For example,
\begin{equation*}
2
\begin{bmatrix}
1 & 2 & 3 \\
4 & 5 & 6
\end{bmatrix} =
\begin{bmatrix}
2 & 4 & 6 \\
8 & 10 & 12
\end{bmatrix} .
\end{equation*}
Matrix addition\index{addition of matrices} is also easy.
We add matrices element by element.
For example,
\begin{equation*}
\begin{bmatrix}
1 & 2 & 3 \\
4 & 5 & 6
\end{bmatrix} +
\begin{bmatrix}
1 & 1 & -1 \\
0 & 2 & 4
\end{bmatrix}
=
\begin{bmatrix}
2 & 3 & 2 \\
4 & 7 & 10
\end{bmatrix} .
\end{equation*}
If the sizes do not match, then addition is not defined.

If we denote by 0 the matrix with all zero entries, by
$c$, $d$ scalars, and by $A$, $B$, $C$ matrices, we
have the following familiar rules:
\begin{align*}
A + 0 & = A = 0 + A , \\
A + B & = B + A , \\
(A + B) + C & = A + (B + C) , \\
c(A+B) & = cA+cB, \\
(c+d)A & = cA + dA.
\end{align*}

Another useful operation for matrices is the so-called
\emph{\myindex{transpose}}.  This operation just swaps rows and columns of a
matrix.
The transpose of $A$ is denoted by $A^T$.  Example:
\begin{equation*}
\begin{bmatrix}
1 & 2 & 3 \\
4 & 5 & 6
\end{bmatrix}^T =
\begin{bmatrix}
1 & 4 \\
2 & 5 \\
3 & 6 
\end{bmatrix}
\end{equation*}

\subsection{Matrix multiplication}

Let us now define matrix multiplication.  First we define the so-called
\emph{\myindex{dot product}} (or \emph{\myindex{inner product}}) of two vectors.
Usually this will be a row vector multiplied
with a column vector of the same size.  For the dot product we multiply
each pair of entries from the first and the second vector and we sum these
products.  The result is a single number.
For example,
\begin{equation*}
\begin{bmatrix}
a_1 & a_2 & a_3
\end{bmatrix}
\cdot
\begin{bmatrix}
b_1 \\
b_2 \\
b_3
\end{bmatrix}
= a_1 b_1 + a_2 b_2 + a_3 b_3 .
\end{equation*}
And similarly for larger (or smaller) vectors.

Armed with the dot product we define the
\emph{\myindex{product of matrices}}.
First let us denote by $\operatorname{row}_i(A)$ the $i^{\text{th}}$ row
of $A$ and by
$\operatorname{column}_j(A)$ the $j^{\text{th}}$ column of $A$.
For an $m \times n$ matrix $A$ and an $n \times p$ matrix $B$
we can define the product $AB$.  We let $AB$ be an $m \times p$
matrix whose $ij^{\text{th}}$ entry is the dot product
\begin{equation*}
\operatorname{row}_i(A) \cdot
\operatorname{column}_j(B) .
\end{equation*}
Do note how the sizes match up: $m \times n$ multiplied by $n \times p$ is 
$m \times p$.  Example:
\begin{multline*}
\begin{bmatrix}
1 & 2 & 3 \\
4 & 5 & 6
\end{bmatrix}
\begin{bmatrix}
1 & 0 & -1 \\
1 & 1 & 1 \\
1 & 0 & 0
\end{bmatrix}
= \\ =
\begin{bmatrix}
1\cdot 1 + 2\cdot 1 + 3 \cdot 1 &  &
1\cdot 0 + 2\cdot 1 + 3 \cdot 0 &  &
1\cdot (-1) + 2\cdot 1 + 3 \cdot 0 \\
4\cdot 1 + 5\cdot 1 + 6 \cdot 1 &  &
4\cdot 0 + 5\cdot 1 + 6 \cdot 0 &  &
4\cdot (-1) + 5\cdot 1 + 6 \cdot 0
\end{bmatrix}
=
\begin{bmatrix}
6 & 2 & 1 \\
15 & 5 & 1
\end{bmatrix}.
\end{multline*}

\medskip

For multiplication we want an analogue of a 1.  This analogue is the
so-called \emph{\myindex{identity matrix}}.
The identity matrix is a square matrix with 1s on the
diagonal and zeros everywhere else.  It is usually denoted by $I$.
For each size we have a different identity matrix and so sometimes we may denote
the size as a subscript.  For example, the $I_3$ would be the $3 \times 3$
identity matrix
\begin{equation*}
I = I_3 =
\begin{bmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{bmatrix} .
\end{equation*}

We have the following rules for matrix multiplication.  Suppose that
$A$, $B$, $C$ are matrices of the correct sizes so that the following
make sense.  Let $\alpha$ denote a scalar (number).
\begin{align*}
A(BC) & = (AB)C, \\
A(B+C) & = AB + AC, \\
(B+C)A & = BA + CA, \\
\alpha(AB) & = (\alpha A)B = A(\alpha B), \\
IA & = A = AI .
\end{align*}

\pagebreak[2]
A few warnings are in order.
\begin{enumerate}[(i)]
\item $AB \not= BA$ in general (it may be true by fluke sometimes).  That is,
matrices do not \myindex{commute}.
For example, take
$A = \left[ \begin{smallmatrix} 1 & 1 \\ 1 & 1 \end{smallmatrix} \right]$
and
$B = \left[ \begin{smallmatrix} 1 & 0 \\ 0 & 2 \end{smallmatrix} \right]$.
\item $AB = AC$ does not necessarily imply $B=C$, even if $A$ is not 0.
\item $AB = 0$ does not necessarily mean that $A=0$ or $B=0$.
Try, for example,
$A = B = \left[ \begin{smallmatrix} 0 & 1 \\ 0 & 0 \end{smallmatrix}
\right]$.
\end{enumerate}

For the last two items to hold we would need to \myquote{divide} by
a matrix.  This is where the \emph{\myindex{matrix inverse}} comes in.
Suppose that $A$ and $B$ are $n \times n$ matrices such that
\begin{equation*}
AB = I = BA .
\end{equation*}
Then we call $B$ the inverse of $A$ and we denote $B$ by $A^{-1}$.
If the inverse of $A$ exists, then we call $A$
\emph{invertible\index{invertible matrix}}.
If $A$ is not invertible, we sometimes say $A$ is
\emph{singular\index{singular matrix}}.

If $A$ is invertible, then $AB = AC$ does imply
that $B = C$ (in particular the inverse of $A$ is unique).
We just multiply both sides by $A^{-1}$ (on the left) to get
$A^{-1}AB = A^{-1}AC$ or $IB=IC$ or $B=C$.
We can also see from the definition that ${(A^{-1})}^{-1} = A$.

\subsection{The determinant}

For square matrices we define a useful quantity called the
\emph{\myindex{determinant}}.  We define
the determinant of a $1 \times 1$ matrix as the value of its only entry.
For a $2 \times 2$ matrix we define
\begin{equation*}
\det \left(
\begin{bmatrix}
a & b \\
c & d
\end{bmatrix}
\right)
\overset{\text{def}}{=}
ad-bc .
\end{equation*}

Before trying to define the
determinant for larger matrices, let us note
the meaning of the determinant.  Consider an $n \times n$ matrix
as a mapping of the $n$-dimensional euclidean space ${\mathbb{R}}^n$ to 
itself, where $\vec{x}$ gets sent to $A \vec{x}$.
In particular, a $2 \times 2$ matrix $A$ is a mapping of
the plane to itself.  The
determinant of 
$A$ is the factor by which the area of objects changes. 
If we take the unit square (square of side 1) in the plane, then
$A$ takes the square to a parallelogram of area $\lvert\det(A)\rvert$.  The sign
of $\det(A)$ denotes changing of orientation (negative if the axes get flipped).  For
example, let
\begin{equation*}
A =
\begin{bmatrix}
1 & 1 \\
-1 & 1
\end{bmatrix} .
\end{equation*}
Then $\det(A) = 1+1 = 2$.  Let us see where the (unit) square with vertices
$(0,0)$, $(1,0)$, $(0,1)$, and $(1,1)$ gets sent.  Clearly $(0,0)$ gets sent
to $(0,0)$.  
\begin{equation*}
\begin{bmatrix}
1 & 1 \\
-1 & 1
\end{bmatrix}
\begin{bmatrix}
1 \\ 0
\end{bmatrix} =
\begin{bmatrix}
1 \\
-1 
\end{bmatrix}
,
\qquad
\begin{bmatrix}
1 & 1 \\
-1 & 1
\end{bmatrix}
\begin{bmatrix}
0 \\ 1
\end{bmatrix} =
\begin{bmatrix}
1 \\
1 
\end{bmatrix}
,
\qquad
\begin{bmatrix}
1 & 1 \\
-1 & 1
\end{bmatrix}
\begin{bmatrix}
1 \\ 1
\end{bmatrix} =
\begin{bmatrix}
2 \\
0 
\end{bmatrix}
.
\end{equation*}
The image of the square is another square with vertices $(0,0)$, $(1,-1)$,
$(1,1)$, and $(2,0)$.  The
image square has
a side of length $\sqrt{2}$ and is therefore of area 2.

If you think back to high school geometry, you may have seen a formula for
computing the area of a \myindex{parallelogram}
with vertices $(0,0)$, $(a,c)$, $(b,d)$
and $(a+b,c+d)$.  And it is precisely
\begin{equation*}
\left\lvert \, \det \left(
\begin{bmatrix} a & b \\ c & d \end{bmatrix}
\right) \, \right\lvert.
\end{equation*}
The vertical lines above mean absolute value.
The matrix $\left[ \begin{smallmatrix} a & b \\ c & d \end{smallmatrix}
\right]$
carries the unit square to the given parallelogram.

\medskip

Let us look at the determinant for larger matrices.  We define $A_{ij}$ as
the matrix $A$ with the $i^{\text{th}}$ row and the $j^{\text{th}}$ column
deleted.  To compute the determinant of a matrix, pick one row, say the
$i^{\text{th}}$ row and compute:
\begin{equation*}
\mybxbg{~~
\det (A) =
\sum_{j=1}^n
{(-1)}^{i+j}
a_{ij} \det (A_{ij}) .
~~}
\end{equation*}
For the first row we get
\begin{equation*}
\det (A) =
a_{11} \det (A_{11}) - 
a_{12} \det (A_{12}) + 
a_{13} \det (A_{13}) - 
\cdots
\begin{cases}
+ a_{1n} \det (A_{1n}) & \text{if } n \text{ is odd,} \\
- a_{1n} \det (A_{1n}) & \text{if } n \text{ even.}
\end{cases}
\end{equation*}
We alternately add and subtract the determinants of the submatrices
$A_{ij}$ multiplied by $a_{ij}$ for a fixed $i$ and all $j$.
For a $3 \times 3$ matrix,
picking the first row, we get $\det (A) = a_{11} \det (A_{11}) -
a_{12} \det (A_{12}) + a_{13} \det (A_{13})$.  For example,
\begin{equation*}
\begin{split}
\det \left(
\begin{bmatrix}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{bmatrix}
\right)
& =
1 \cdot
\det \left(
\begin{bmatrix}
5 & 6 \\
8 & 9
\end{bmatrix}
\right)
-
2 \cdot
\det \left(
\begin{bmatrix}
4 & 6 \\
7 & 9
\end{bmatrix}
\right)
+
3 \cdot
\det \left(
\begin{bmatrix}
4 & 5 \\
7 & 8
\end{bmatrix}
\right) \\
& =
1 (5 \cdot 9 - 6 \cdot 8)
-
2 (4 \cdot 9 - 6 \cdot 7)
+
3 (4 \cdot 8 - 5 \cdot 7)
= 0 .
\end{split}
\end{equation*}

The numbers ${(-1)}^{i+j}\det(A_{ij})$ are called
\emph{cofactors\index{cofactor}}
of the matrix and
this way of computing the determinant is called the
\emph{\myindex{cofactor expansion}}.
No matter which row you pick, you always get the same number.
It is also possible to compute the
determinant by expanding
along columns (picking a column instead of a row above).
It is true that $\det(A) = \det(A^T)$.

A common notation for the determinant is a pair of vertical
lines:
\begin{equation*}
\begin{vmatrix}
a & b \\
c & d
\end{vmatrix}
=
\det \left(
\begin{bmatrix}
a & b \\
c & d
\end{bmatrix}
\right) .
\end{equation*}
I personally find this notation confusing as vertical lines usually
mean a positive quantity, while determinants can be negative.  Also
think about how to write the absolute value of a determinant.  I will not
use this notation in this book.

\medskip

Think of the determinants telling you the scaling of a mapping.  
If $B$ doubles the sizes of geometric objects and $A$ triples them,
then $AB$ (which applies $B$ to an object and then $A$) should make size
go up by a factor of $6$.  This is true in general:
\begin{equation*}
\det(AB) = \det(A)\det(B) .
\end{equation*}
This property is one of the most useful, and it is employed often to 
actually compute determinants.  A particularly interesting consequence is to
note what it means for existence of inverses.
Take $A$ and $B$ to be inverses of each other, that is $AB=I$.  Then
\begin{equation*}
\det(A)\det(B) = \det(AB) = \det(I) = 1 .
\end{equation*}
Neither $\det(A)$ nor $\det(B)$ can be zero.
Let us state this as a theorem
as it will be very important in the context of this course.

\begin{theorem1}{}
An $n \times n$ matrix $A$ is invertible if and only if $\det (A) \not= 0$.
\end{theorem1}

In fact, $\det(A^{-1}) \det(A) = 1$ says that $\det(A^{-1}) =
\frac{1}{\det(A)}$.  So we even know what the determinant of $A^{-1}$ is
before we know how to compute $A^{-1}$.

There is a simple formula for the inverse of a $2 \times 2$ matrix
\begin{equation*}
\begin{bmatrix}
a & b \\
c & d
\end{bmatrix}^{-1}
=
\frac{1}{ad-bc}
\begin{bmatrix}
d & -b \\
-c & a
\end{bmatrix} .
\end{equation*}
Notice the determinant of the matrix
$[\begin{smallmatrix}a&b\\c&d\end{smallmatrix}]$
in the denominator of the fraction.
The formula only works if the determinant is nonzero, otherwise we are
dividing by zero.

\subsection{Solving linear systems}

One application of matrices we will need is to solve systems of
linear equations.  This is best shown by example.
Suppose that we have the following system of linear equations
\begin{align*}
          2 x_1 +           2 x_2 +           2 x_3 & = 2 , \\
\phantom{9} x_1 + \phantom{9} x_2 +           3 x_3 & = 5 , \\
\phantom{9} x_1 +           4 x_2 + \phantom{9} x_3 & = 10 .
\end{align*}

Without changing the solution,
we could swap equations in this system,
we could multiply any of the equations by a nonzero number, and
we could add a multiple of one equation to another equation.
It turns out these operations always suffice to find a solution.

It is easier to write the system as a matrix equation.
The system above can be
written as
\begin{equation*}
\begin{bmatrix}
2 & 2 & 2 \\
1 & 1 & 3 \\
1 & 4 & 1 
\end{bmatrix}
\begin{bmatrix}
x_1 \\
x_2 \\
x_3
\end{bmatrix} 
=
\begin{bmatrix}
2 \\
5 \\
10
\end{bmatrix} .
\end{equation*}
To solve the system we put the coefficient matrix (the matrix on the
left-hand side of the equation) together with the vector on the right and side
and get the
so-called
\emph{\myindex{augmented matrix}}
\begin{equation*}
\left[
\begin{array}{ccc|c}
2 & 2 & 2 & 2 \\
1 & 1 & 3 & 5 \\
1 & 4 & 1 & 10
\end{array}
\right] .
%\qquad
%\text{or just}
%\qquad
%\begin{bmatrix}
%2 & 2 & 2 & 2 \\
%1 & 1 & 3 & 5 \\
%1 & 4 & 1 & 10
%\end{bmatrix} .
\end{equation*}

\pagebreak[2]
We apply the following three elementary operations.
\begin{enumerate}[(i)]
\item Swap two rows.
\item Multiply a row by a nonzero number.
\item Add a multiple of one row to another row.
\end{enumerate}
We keep doing these operations until we get into a state where it is
easy to read off the answer, or until we get into a contradiction indicating
no solution, for example if we come up with an equation such as $0=1$.

Let us work through the example.  First multiply the first row by
$\nicefrac{1}{2}$ to obtain
\begin{equation*}
\left[
\begin{array}{ccc|c}
1 & 1 & 1 & 1 \\
1 & 1 & 3 & 5 \\
1 & 4 & 1 & 10
\end{array}
\right] .
\end{equation*}
Now subtract the first row from the second and third row.
\begin{equation*}
\left[
\begin{array}{ccc|c}
1 & 1 & 1 & 1 \\
0 & 0 & 2 & 4 \\
0 & 3 & 0 & 9
\end{array}
\right]
\end{equation*}
Multiply the last row by $\nicefrac{1}{3}$ and the second row by $\nicefrac{1}{2}$.
\begin{equation*}
\left[
\begin{array}{ccc|c}
1 & 1 & 1 & 1 \\
0 & 0 & 1 & 2 \\
0 & 1 & 0 & 3
\end{array}
\right]
\end{equation*}
Swap rows 2 and 3.
\begin{equation*}
\left[
\begin{array}{ccc|c}
1 & 1 & 1 & 1 \\
0 & 1 & 0 & 3 \\
0 & 0 & 1 & 2
\end{array}
\right]
\end{equation*}
Subtract the last row from the first, then subtract the second row
from the first.
\begin{equation*}
\left[
\begin{array}{ccc|c}
1 & 0 & 0 & -4 \\
0 & 1 & 0 & 3 \\
0 & 0 & 1 & 2
\end{array}
\right]
\end{equation*}
If we think about what equations this augmented matrix represents, we see that
$x_1 = -4$, $x_2 = 3$, and $x_3 = 2$.  We try this solution in the original
system and, voil\`a, it works!

\begin{exercise}
Check that the solution above really solves the given equations.
\end{exercise}

We write this equation in matrix notation as
\begin{equation*}
A \vec{x} = \vec{b} ,
\end{equation*}
where $A$ is the matrix
$\left[ \begin{smallmatrix}
2 & 2 & 2 \\
1 & 1 & 3 \\
1 & 4 & 1 
\end{smallmatrix} \right]$ and $\vec{b}$ is the vector
$\left[ \begin{smallmatrix}
2 \\
5 \\
10
\end{smallmatrix} \right]$.  The solution can also be computed via the
inverse,
\begin{equation*}
\vec{x} = A^{-1} A \vec{x} = A^{-1} \vec{b} .
\end{equation*}

\medskip

It is
possible that the solution is not unique, or that no solution exists.
It is easy to tell if a solution does not exist.  If during the row
reduction you come up with a row where all the entries except the last one
are zero (the last entry in a row corresponds to the right-hand side of the
equation), then the system is \emph{inconsistent\index{inconsistent system}} and
has no solution.  For
example, for a system of 3 equations and 3 unknowns, if you find a row
such as $[\,0 \quad 0 \quad 0 ~\,|\,~ 1\,]$ in the augmented matrix,
you know the system is inconsistent.  That row corresponds to $0=1$.

\medskip

You generally try to use row operations until the following conditions
are satisfied.  The first (from the left) nonzero entry in each row is called the
\emph{\myindex{leading entry}}.
\begin{enumerate}[(i)]
\item The leading entry in any row is strictly to the right of
the leading entry of the row above.
\item Any zero rows are below all the nonzero rows.
\item All leading entries are 1.
\item All the entries above and below a leading entry are zero.
\end{enumerate}
Such a matrix is said to be in
\emph{\myindex{reduced row echelon form}}.  The variables
corresponding to columns with no leading entries are said to be
\emph{free variables\index{free variable}}.
Free variables mean that we can pick those variables
to be anything we want and then solve for the rest of the unknowns.

\begin{example}
The following augmented matrix is in reduced row echelon form.
\begin{equation*}
\left[
\begin{array}{ccc|c}
1 & 2 & 0 & 3 \\
0 & 0 & 1 & 1 \\
0 & 0 & 0 & 0
\end{array}
\right]
\end{equation*}
Suppose
the variables are $x_1$, $x_2$, and $x_3$.  Then $x_2$ is the
free variable, $x_1 = 3 - 2x_2$, and $x_3 = 1$.

\medskip

On the other hand if during the row reduction process you come up with the
matrix
\begin{equation*}
\left[
\begin{array}{ccc|c}
1 & 2 & 13 & 3 \\
0 & 0 & 1 & 1 \\
0 & 0 & 0 & 3
\end{array}
\right]
,
\end{equation*}
there is no need to go further.  The last row corresponds to
the equation $0 x_1 + 0 x_2 + 0 x_3 = 3$, which is preposterous.  Hence, no
solution exists.
\end{example}

\subsection{Computing the inverse}

If the matrix $A$ is square and there exists a unique solution
$\vec{x}$ to $A \vec{x} = \vec{b}$ for any $\vec{b}$ (there are no free
variables), then $A$ is invertible.
Multiplying both sides by $A^{-1}$, you can see that $\vec{x} =
A^{-1} \vec{b}$.  So it is useful to compute the inverse if you want to
solve the equation for many different right-hand sides $\vec{b}$.

We have a formula for
the $2 \times 2$ inverse, but it is also not hard
to compute inverses of larger matrices.
While we will not have too much occasion to compute inverses for larger
matrices than $2 \times 2$ by hand, let us touch on how to do it.
Finding the inverse of $A$ is actually just solving a bunch of linear
equations.  If we can solve $A \vec{x}_k = \vec{e}_k$ where $\vec{e}_k$ is
the vector with all zeros except a 1 at the $k^{\text{th}}$ position, then
the inverse is the matrix with the columns $\vec{x}_k$ for $k=1,2,\ldots,n$
(exercise: why?).  Therefore, to find the inverse we write a larger $n
\times 2n$ augmented matrix $[ \,A ~|~ I\, ]$, where $I$ is the identity
matrix.
We then perform row reduction.
The reduced row echelon form of $[ \,A ~|~ I\, ]$ 
will be of the form $[ \,I ~|~ A^{-1}\, ]$ if and only if
$A$ is invertible.  We then just read off the inverse $A^{-1}$.

\subsection{Eigenvalues and eigenvectors of a matrix}

Let $A$ be a constant square matrix.  Suppose there is a
scalar $\lambda$ and a nonzero vector $\vec{v}$ such that
\begin{equation*}
A \vec{v} = \lambda \vec{v}.
\end{equation*}
We call $\lambda$ an \emph{\myindex{eigenvalue}} of $A$ and we
call $\vec{v}$
a corresponding \emph{\myindex{eigenvector}}.

\begin{example}
The matrix $\left[ \begin{smallmatrix}
2 & 1 \\
0 & 1
\end{smallmatrix} \right]$ has an eigenvalue $\lambda = 2$ with a
corresponding
eigenvector $\left[ \begin{smallmatrix}
1 \\ 0
\end{smallmatrix} \right]$ as
\begin{equation*}
\begin{bmatrix}
2 & 1 \\
0 & 1
\end{bmatrix}
\begin{bmatrix}
1 \\ 0
\end{bmatrix}
=
\begin{bmatrix}
2 \\
0 
\end{bmatrix}
=
2
\begin{bmatrix}
1 \\ 0
\end{bmatrix} .
\end{equation*}
\end{example}

Let us see how to compute eigenvalues for any matrix.
Rewrite the equation for an eigenvalue as
\begin{equation*}
(A - \lambda I)\vec{v} = \vec{0} .
\end{equation*}
This equation has a nonzero solution $\vec{v}$ only if 
$A - \lambda I$ is not invertible.  Were it invertible,
we could write
${(A - \lambda I)}^{-1}(A - \lambda I)\vec{v} = {(A-\lambda I)}^{-1}\vec{0}$,
which implies $\vec{v} = \vec{0}$.  Therefore,
$A$ has the
eigenvalue $\lambda$ if and only if $\lambda$ solves the equation
\begin{equation*}
\det (A-\lambda I) = 0 .
\end{equation*}

Consequently, we will be able to find an eigenvalue of $A$ without
finding a corresponding eigenvector.  An eigenvector will have to be
found later, once $\lambda$ is known.

\begin{example}
Find all eigenvalues of 
$\left[ \begin{smallmatrix}
2 & 1 & 1 \\
1 & 2 & 0 \\
0 & 0 & 2
\end{smallmatrix} \right]$.
\end{example}

\begin{exampleSol}
We write
\begin{multline*}
\det \left(
\begin{bmatrix}
2 & 1 & 1 \\
1 & 2 & 0 \\
0 & 0 & 2
\end{bmatrix}
- \lambda 
\begin{bmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{bmatrix}
\right)
=
\det \left(
\begin{bmatrix}
2-\lambda & 1 & 1 \\
1 & 2-\lambda & 0 \\
0 & 0 & 2-\lambda
\end{bmatrix}
\right)
= \\
=
(2-\lambda) \bigl({(2-\lambda)}^2 - 1\bigr)
= 
-(\lambda -1)(\lambda -2)(\lambda-3) .
\end{multline*}
So the eigenvalues are $\lambda = 1$, $\lambda = 2$, and
$\lambda = 3$.
\end{exampleSol}

For an $n \times n$ matrix, the polynomial we get by
computing $\det(A - \lambda I)$ is of degree $n$, and hence
in general, we have $n$ eigenvalues.  Some may be repeated, some may be
complex.

\medskip

To find an eigenvector corresponding to an eigenvalue $\lambda$, we write
\begin{equation*}
(A-\lambda I) \vec{v} = \vec{0} ,
\end{equation*}
and solve for a nontrivial (nonzero) vector $\vec{v}$.
If $\lambda$ is an eigenvalue, there will be at least one
free variable, and so
for
each distinct eigenvalue $\lambda$, we can always find an eigenvector.

\begin{example}
Find an eigenvector of
$\left[ \begin{smallmatrix}
2 & 1 & 1 \\
1 & 2 & 0 \\
0 & 0 & 2
\end{smallmatrix} \right]$ corresponding to the eigenvalue $\lambda = 3$.
\end{example}

\begin{exampleSol}
We write
\begin{equation*}
(A-\lambda I) \vec{v} = 
\left(
\begin{bmatrix}
2 & 1 & 1 \\
1 & 2 & 0 \\
0 & 0 & 2
\end{bmatrix}
- 3
\begin{bmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{bmatrix}
\right)
\begin{bmatrix}
v_1 \\ v_2 \\ v_3
\end{bmatrix}
=
\begin{bmatrix}
-1 & 1 & 1 \\
1 & -1 & 0 \\
0 & 0 & -1
\end{bmatrix}
\begin{bmatrix}
v_1 \\ v_2 \\ v_3
\end{bmatrix}
=
\vec{0} .
\end{equation*}
It is easy to solve this system of linear equations.  We write down the
augmented matrix
\begin{equation*}
\left[
\begin{array}{ccc|c}
-1 & 1 & 1 & 0 \\
1 & -1 & 0 & 0 \\
0 & 0 & -1 & 0
\end{array}
\right] ,
\end{equation*}
and perform row operations (exercise: which ones?) until we get:
\begin{equation*}
\left[
\begin{array}{ccc|c}
1 & -1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 0
\end{array}
\right] .
\end{equation*}
The entries of $\vec{v}$ have to satisfy the equations
$v_1 - v_2 = 0$, $v_3 = 0$, and $v_2$ is a free variable.  
We can pick $v_2$ to be arbitrary (but nonzero), let
$v_1 = v_2$, and of course $v_3 = 0$.
For example, if we pick $v_2 = 1$, then
$\vec{v} =
\left[ \begin{smallmatrix} 1 \\ 1 \\ 0 \end{smallmatrix} \right]$.
Let us verify that $\vec{v}$ really is an eigenvector corresponding to $\lambda = 3$:
\begin{equation*}
\begin{bmatrix}
2 & 1 & 1 \\
1 & 2 & 0 \\
0 & 0 & 2 \\
\end{bmatrix}
\begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix}
=
\begin{bmatrix}
3 \\
3 \\
0
\end{bmatrix}
=
3
\begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} .
\end{equation*}
Yay!  It worked.
\end{exampleSol}

\begin{exercise}[easy]
Are eigenvectors unique?  Can you find a different eigenvector for
$\lambda = 3$ in the example above?  How are the two eigenvectors related?
\end{exercise}

\begin{exercise}
When the matrix is $2 \times 2$ you do not need to
%write down the augmented matrix and
do row operations when computing an eigenvector,
you can read it off from $A-\lambda I$
(if you have computed the eigenvalues correctly).
Can you see why?  Explain.  Try it for the matrix
$\left[ \begin{smallmatrix} 2 & 1 \\ 1 & 2 \end{smallmatrix} \right]$.
\end{exercise}


\subsection{Exercises}

\begin{exercise}
Let $A$ and $B$ be the matrices below.
\[ A = \begin{bmatrix} 1 & 4 & -1 \\ 2 & 0 & 3 \\
1 & -2 & 3 \end{bmatrix} \qquad B = \begin{bmatrix} 0 & 2 & 3 \\ 1 & -4 & -2 \\ 2 & -5 & 1 \end{bmatrix} \]
Compute $A + 3B$, $AB$, and $BA$. 
\end{exercise}

\begin{exercise}
Solve
$\left[ \begin{smallmatrix}
1 & 2 \\
3 & 4 
\end{smallmatrix} \right] \vec{x} =
\left[ \begin{smallmatrix}
5 \\
6
\end{smallmatrix} \right]$ by using matrix inverse.
\end{exercise}

\begin{exercise}
Compute determinant of
$\left[ \begin{smallmatrix}
9 & -2 & -6 \\
-8 & 3 & 6 \\
10 & -2 & -6
\end{smallmatrix} \right]$.
\end{exercise}

\begin{exercise}\ansMark%
Compute determinant of
$\left[ \begin{smallmatrix}
1 & 1 & 1 \\
2 & 3 & -5 \\
1 & -1 & 0
\end{smallmatrix}\right]$
\end{exercise}
\exsol{%
$-15$
}

\begin{exercise}
Compute determinant of
$\left[ \begin{smallmatrix}
1 & 2 & 3 & 1 \\
4 & 0 & 5 & 0 \\
6 & 0 & 7 & 0 \\
8 & 0 & 10 & 1
\end{smallmatrix} \right]$.  Hint: Expand along the proper row or column to
make the calculations simpler.
\end{exercise}

\begin{exercise}
Compute inverse of
$\left[ \begin{smallmatrix}
1 & 2 & 3 \\
1 & 1 & 1 \\
0 & 1 & 0
\end{smallmatrix} \right]$.
\end{exercise}

\begin{exercise}
For which $h$ is
$\left[ \begin{smallmatrix}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & h
\end{smallmatrix} \right]$
not invertible?  Is there only one such $h$?  Are there several?  Infinitely
many?
\end{exercise}


\begin{exercise}\ansMark%
Find $t$ such that
$\left[ \begin{smallmatrix}
1 & t \\
-1 & 2
\end{smallmatrix}\right]$
is not invertible.
\end{exercise}
\exsol{%
$-2$
}

\begin{exercise}
For which $h$ is
$\left[ \begin{smallmatrix}
h & 1 & 1 \\
0 & h & 0 \\
1 & 1 & h
\end{smallmatrix} \right]$
not invertible?  Find all such $h$.
\end{exercise}

\begin{exercise}\ansMark%
Solve the system of equations
\begin{equation*}
\begin{split}
 4x_1 - 2x_2 + 4x_3 &= -8 \\ 
 x_1 - 3x_3 &= 12\\
-4x_1 + 4x_2 + 4x_3 &= -8 
\end{split}
\end{equation*}
or determine that no solution exists.
\end{exercise}
\exsol{%
$x_1 = 3$, $x_2 = 4$, $x_3 = -3$.
}%

\begin{exercise}\ansMark%
Solve the system of equations
\begin{equation*}
\begin{split}
 -x_1 - 4x_2 + 2x_3 &= 11 \\ 
 3x_1 -3x_2 + x_3 &= 13\\
-5x_1 - 5x_2 + 3x_3 &= 9 
\end{split}
\end{equation*}
or determine that no solution exists.
\end{exercise}
\exsol{%
Infinitely many solutions of the form $x_1 = \frac{19}{15} + \frac{2}{15}t$, $x_2 = \frac{7}{15}t - \frac{46}{15}$, $x_3 = t$ for any real number $t$. 
}%

\begin{exercise}\ansMark%
Solve the system of equations
\begin{equation*}
\begin{split}
x_1 + 3x_2 - 3x_3 &= 1 \\ 
 -3x_1 - 4x_2  +4x_3 &= -3\\
4x_1 + 7x_2 - 7x_3 &= 7 
\end{split}
\end{equation*}
or determine that no solution exists.
\end{exercise}
\exsol{%
No solution.
}%

\begin{exercise}\ansMark%
Solve the system of equations
\begin{equation*}
\begin{split}
 x_1  + 3x_2  - x_3 &= 5 \\ 
 2x_1 +x_2 &= -3\\
-3x_1 - 4x_2 + 2x_3 &= -6 
\end{split}
\end{equation*}
or determine that no solution exists.
\end{exercise}
\exsol{%
$x_1 = -2$, $x_2 = 1$, $x_3 = -4$.
}%

\begin{exercise}
Solve
$\left[ \begin{smallmatrix}
9 & -2 & -6 \\
-8 & 3 & 6 \\
10 & -2 & -6
\end{smallmatrix} \right] \vec{x} =
\left[ \begin{smallmatrix}
1 \\
2 \\
3
\end{smallmatrix} \right]$ .
\end{exercise}

\begin{exercise}
Solve
$\left[ \begin{smallmatrix}
5 & 3 & 7 \\
8 & 4 & 4 \\
6 & 3 & 3
\end{smallmatrix} \right] \vec{x} =
\left[ \begin{smallmatrix}
2 \\
0 \\
0
\end{smallmatrix} \right]$.
\end{exercise}

\begin{exercise}\ansMark%
Solve
$\left[ \begin{smallmatrix}
1 & 1 \\
1 & -1
\end{smallmatrix}\right] \vec{x} = 
\left[ \begin{smallmatrix}
10 \\ 20
\end{smallmatrix}\right]$.
\end{exercise}
\exsol{%
$\vec{x} =
\left[ \begin{smallmatrix}
15 \\ -5
\end{smallmatrix}\right]$
}


\begin{exercise}
Solve
$\left[ \begin{smallmatrix}
3 & 2 & 3 & 0 \\
3 & 3 & 3 & 3 \\
0 & 2 & 4 & 2 \\
2 & 3 & 4 & 3 
\end{smallmatrix} \right] \vec{x} =
\left[ \begin{smallmatrix}
2 \\
0 \\
4 \\
1
\end{smallmatrix} \right]$.
\end{exercise}

\begin{exercise}
Find 3 nonzero $2 \times 2$ matrices $A$, $B$, and $C$ such that
$AB = AC$ but $B \not= C$.
\end{exercise}

\begin{exercise}\ansMark%
Suppose $a, b, c$ are nonzero numbers.
Let
$M=\left[ \begin{smallmatrix}
a & 0 \\
0 & b
\end{smallmatrix}\right]$,
$N=\left[ \begin{smallmatrix}
a & 0 & 0 \\
0 & b & 0 \\
0 & 0 & c
\end{smallmatrix}\right]$.
\begin{tasks}(2)
\task Compute $M^{-1}$.
\task Compute $N^{-1}$.
\end{tasks}
\end{exercise}
\exsol{%
a) $\left[ \begin{smallmatrix}
\nicefrac{1}{a} & 0 \\
0 & \nicefrac{1}{b}
\end{smallmatrix}\right]$
\quad
b)
$\left[ \begin{smallmatrix}
\nicefrac{1}{a} & 0 & 0 \\
0 & \nicefrac{1}{b} & 0 \\
0 & 0 & \nicefrac{1}{c}
\end{smallmatrix}\right]$
}

\begin{exercise}[easy]
Let $A$ be a $3 \times 3$ matrix with an eigenvalue of 3 and a
corresponding eigenvector $\vec{v} =
\left[ \begin{smallmatrix} 1 \\ -1 \\ 3 \end{smallmatrix} \right]$.
Find $A \vec{v}$.
\end{exercise}

\begin{exercise}\ansMark%
Find the eigenvalues and eigenvectors for the matrix
\[ \begin{bmatrix} 0 & -2 \\ 1 & 3\end{bmatrix}. \]
\end{exercise}
\exsol{%
$\lambda_1 = 1$, $\vec{v}_1 = \begin{bmatrix} -2 \\ 1\end{bmatrix}$, $\lambda_2 = 2$, $\vec{v}_2 = \begin{bmatrix} 1 \\ -1 \end{bmatrix}$. 
}

\begin{exercise}\ansMark%
Find the eigenvalues and eigenvectors for the matrix
\[ \begin{bmatrix} -8 & -5 \\ 8 & 4\end{bmatrix}. \]
\end{exercise}
\exsol{%
$\lambda_1 = -2 + 2i$, $\vec{v}_1 = \begin{bmatrix} -3+i \\ 4\end{bmatrix}$, $\lambda_2 = -2-2i$, $\vec{v}_2 = \begin{bmatrix} -3-i \\ 4 \end{bmatrix}$. 
}

\begin{exercise}\ansMark%
Find the eigenvalues and eigenvectors for the matrix
\[ \begin{bmatrix} 7 & -3 & 7 \\ 9 & -5 & 7 \\ 0 & 0 & -3\end{bmatrix}. \]
\end{exercise}
\exsol{%
$\lambda_1 = 4$, $\vec{v}_1 = \begin{bmatrix} 1 \\ 1 \\ 0\end{bmatrix}$, $\lambda_2 = -2$, $\vec{v}_2 = \begin{bmatrix} 1 \\ 3 \\ 0 \end{bmatrix}$, $\lambda_3 = -3$, $\vec{v}_3 = \begin{bmatrix} -1 \\ -1 \\ 1 \end{bmatrix}$. 
}

\setcounter{exercise}{100}
